---
title: "Lecture 9: Student Notes"
author: "FBA Instructors"
format: html
---

## Learning Goals

By the end of this lecture, you should be able to:

* explain why transaction-level data often need to be aggregated to the customer level for marketing decisions
* define **Recency**, **Frequency**, and **Monetary value** in the context of customer analytics
* construct customer-level RFM measures from transaction data
* use RFM measures to create and interpret rule-based customer segments
* explain why scaling is useful before clustering
* interpret k-means clusters as a data-driven alternative to rule-based segmentation
* compare the strengths and limitations of rule-based and data-driven segmentation

## The Business Challenge: From Transactions to Customers

Suppose you are a marketing analyst working with data from the **CDNOW Online Music Store**, a retailer from the early years of e-commerce. Your dataset contains detailed information on customers’ purchases between 1997 and 1998, including:

* who bought
* when they bought
* how much they spent

Your manager asks a practical question:

> Which customers should we target for our new loyalty campaign?

This is a customer segmentation problem. The company does not want to treat every customer in the same way. Some customers buy often, some spend more, and some may be drifting away. Treating every customer the same can waste marketing resources and miss opportunities.

A key challenge is that the data are currently stored at the **transaction level**, while the manager’s question is about **customers**. A long table of purchases tells us what happened in each transaction, but it does not directly tell us what kind of customer each person is.

This is where descriptive analytics becomes useful. **Descriptive analytics is not just a summary of the past, but also a way to organise the past into forms that are useful for business action.** In this lecture, we will turn transaction histories into customer-level summaries and then use those summaries to build customer segments.

We will compare two approaches:

1. **RFM segmentation**, a simple rule-based approach used widely in practice
2. **K-means clustering**, a data-driven approach that lets the data suggest groups of similar customers

## RFM Segmentation in Practice

One common way to move from raw purchase data to useful customer segments is **RFM analysis**. In marketing and customer relationship management, **RFM** stands for:

* **Recency**: how recently a customer made a purchase
* **Frequency**: how often a customer has purchased
* **Monetary value**: how much a customer has spent

These three measures are widely used because they are simple, interpretable, and often useful for business action.

For example:

* a customer who purchased recently, purchases often, and spends a lot may be valuable to retain
* a customer who used to buy regularly but has not purchased for a while may be at risk of churn
* a customer who has purchased only once may need a different strategy from a loyal repeat buyer

RFM is therefore a practical way to classify customers based on their purchase patterns. It is a **rule-based** method because the analyst decides how to convert the RFM measures into segment labels.

Before we can segment customers, we first need to transform the raw transaction data into customer-level summaries.

## Building Rule-Based Customer Segments

### Load packages and data

We will use the following packages:

```{r}
library(tidyverse)   # data manipulation and visualisation
library(lubridate)   # working with dates
library(recipes)     # preprocessing pipelines
library(factoextra)  # visualising k-means clusters
library(broom)       # attaching model output back to data
```

Now load the CDNOW transaction data:

```{r}
purchases <- read_csv("../data/cdnow_purchases.csv")

glimpse(purchases)
```

Each row in this dataset is a **transaction**, not a customer. That means our first task is to aggregate these purchases to the customer level.

### Step 1: Aggregate transaction data to the customer level

To compute RFM measures, we need one row per customer. We begin by calculating:

* each customer’s most recent purchase date
* the number of purchases they made
* the total amount they spent

```{r}
customer_aggregates <-
  purchases |>
  group_by(users_id) |>
  summarise(
    last_purch_date = max(date),
    frequency       = n(),
    monetary        = sum(amt),
    .groups = "drop"
  )
```

At this stage, we have started to move from **transactions** to **customers**.

### Step 2: Compute recency

To calculate recency, we need a reference date. Since these data come from 1997 to 1998, it would not make sense to calculate recency relative to today. Instead, we will imagine that we are doing this analysis on **1 July 1998**.

```{r}
analysis_date <- as_date("1998-07-01")
```

Now we compute recency as the number of months between a customer’s last purchase and the analysis date.

```{r}
customer_aggregates <-
  customer_aggregates |>
  mutate(
    recency = interval(last_purch_date, analysis_date) %/% months(1)
  ) |>
  select(-last_purch_date)

glimpse(customer_aggregates)
```

Now each row represents a **customer**, described by three useful behavioural summaries:

* `recency`
* `frequency`
* `monetary`

### Step 3: Convert RFM measures into scores

Raw RFM values are useful, but firms often convert them into ranked scores so that customers can be compared more easily. One simple approach is to divide each variable into five groups using `ntile()`.

```{r}
rfm_data <-
  customer_aggregates |>
  mutate(
    r_tile = ntile(recency, 5),
    f_tile = ntile(frequency, 5),
    m_tile = ntile(monetary, 5)
  )
```

There is one detail to fix. For `recency`, smaller values mean a customer purchased **more recently**, but we often want higher scores to mean “better” or “more engaged.” So we flip the recency score.

```{r}
rfm_data <-
  rfm_data |>
  mutate(
    r_tile = max(r_tile) - r_tile + 1
  )

rfm_data |>
  group_by(r_tile) |>
  summarise(
    mean_r = mean(recency),
    .groups = "drop"
  )
```

After this change, a higher `r_tile` means a more recent customer.

### Step 4: Assign rule-based customer segments

In practice, firms often use predefined business rules to assign customers to named groups. The code below uses a `case_when()` statement to classify customers based on their RFM scores.

You do **not** need to memorise this whole classification system. The key idea is that businesses often turn RFM scores into human-readable labels that support action.

```{r}
rfm_segmented <-
  rfm_data |>
  mutate(
    segment = case_when(
      r_tile == 5 & f_tile == 5 & m_tile == 5 ~ "Champions",
      r_tile %in% 3:5 & f_tile %in% 3:5 & m_tile %in% 2:5 ~ "Potential Loyalist",
      r_tile %in% 2:4 & f_tile %in% 2:4 & m_tile %in% 2:4 ~ "Loyal Customers",
      r_tile %in% 3:4 & f_tile %in% 1:3 & m_tile %in% 3:5 ~ "Promising",
      r_tile %in% 4:5 & f_tile %in% 1:3 & m_tile %in% 1:5 ~ "New Customers",
      r_tile %in% 1:2 & f_tile %in% 3:4 & m_tile %in% 4:5 ~ "Can't Lose Them",
      r_tile %in% 1:2 & f_tile %in% 2:5 & m_tile %in% 4:5 ~ "At Risk",
      r_tile %in% 1:3 & f_tile %in% 3:5 & m_tile %in% 3:5 ~ "Need Attention",
      r_tile %in% 2:3 & f_tile %in% 1:3 & m_tile %in% 1:4 ~ "About To Sleep",
      r_tile == 1 & f_tile %in% 1:5 & m_tile %in% 1:5 ~ "Lost",
      .default = "Unclassified"
    )
  )
```

### Step 5: Summarise the segments

Once the segments are defined, we can summarise them to understand what kinds of customers each group contains.

```{r}
rfm_segmented |>
  group_by(segment) |>
  summarise(
    avg_recency   = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary  = mean(monetary),
    n_customers   = n(),
    .groups = "drop"
  ) |>
  arrange(desc(n_customers)) |>
  kableExtra::kable(digits = 2)
```

This is the main strength of rule-based segmentation: the groups are easy to explain and easy to communicate. The trade-off is that the thresholds are chosen by analysts rather than discovered from the data itself.

## A Data-Driven Alternative: Clustering Customers

Rule-based segmentation is simple and interpretable, but it also depends on cutoffs chosen by analysts. A data-driven alternative is to let an algorithm search for groups of customers who behave similarly.

One widely used method is **k-means clustering**.

### Step 1: Check variable ranges

Before using k-means, it is important to remember that our three variables are measured on different scales.

```{r}
rfm_data |>
  summarise(
    min_recency   = min(recency),
    max_recency   = max(recency),
    min_frequency = min(frequency),
    max_frequency = max(frequency),
    min_monetary  = min(monetary),
    max_monetary  = max(monetary)
  ) |>
  pivot_longer(
    everything(),
    names_to = c("stat", "variable"),
    names_sep = "_"
  ) |>
  pivot_wider(
    names_from = stat,
    values_from = value
  )
```

Because k-means is based on distances, variables with larger numerical scales can dominate the clustering if we do not standardise them first.

### Step 2: Scale the data with `recipes`

A convenient way to preprocess data is to define a **recipe**. Here, we normalise the three numeric predictors so that each has mean 0 and standard deviation 1.

```{r}
rfm_rec <-
  recipe(~ recency + frequency + monetary, data = rfm_data) |>
  step_normalize(all_numeric_predictors())

rfm_prepped <- prep(rfm_rec)
rfm_scaled  <- bake(rfm_prepped, new_data = NULL)

head(rfm_scaled)
```

We can verify that the scaling worked as intended.

```{r}
rfm_scaled |>
  summarise(
    across(everything(), list(mean = mean, sd = sd))
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) |>
  kableExtra::kable(digits = 2)
```

### Step 3: Fit a k-means model

We now apply k-means clustering to the scaled RFM data. For now, we will set the number of clusters to **3**.

```{r}
set.seed(123)

kmeans_3 <- kmeans(rfm_scaled, centers = 3)
```

### Step 4: Visualise the clusters

```{r}
fviz_cluster(
  kmeans_3,
  data = rfm_scaled,
  geom = "point",
  ggtheme = theme_minimal()
)
```

This plot helps us see whether the algorithm has separated customers into distinct groups based on their RFM behaviour.

### Step 5: Attach cluster labels back to the data

The clustering model gives each customer a cluster label. We now attach those labels back to the original customer-level data so that we can interpret the clusters.

```{r}
rfm_clustered <-
  augment(kmeans_3, rfm_data)

head(rfm_clustered)
```

At this point, it is useful to summarise the clusters in the same way we summarised the RFM segments.

```{r}
rfm_clustered |>
  group_by(.cluster) |>
  summarise(
    avg_recency   = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary  = mean(monetary),
    n_customers   = n(),
    .groups = "drop"
  ) |>
  arrange(.cluster) |>
  kableExtra::kable(digits = 2)
```

The interpretation step matters. Clusters do not come with names automatically. We need to examine their average RFM profiles and then decide what kinds of customers they appear to represent.

### If time: how many clusters should we use?

One common way to think about the choice of `k` is to compare within-cluster variation for different numbers of clusters.

```{r}
set.seed(42)

rfm_scale_sample <-
  rfm_scaled |>
  sample_n(1000)
```

```{r}
fviz_nbclust(
  rfm_scale_sample,
  kmeans,
  method = "wss",
  k.max = 15,
  nstart = 5
)
```

This is an extension question for today. The key idea is that choosing `k` is partly an analytical judgment rather than something the algorithm decides automatically.

## Comparing the Two Approaches

In this lecture, we used the same purchase histories to build customer segments in two different ways.

**RFM segmentation** is a **rule-based** approach. It is useful because it is easy to compute, easy to explain, and closely aligned with business action. Managers can understand segment labels such as *Champions* or *At Risk* and connect them to different marketing strategies.

**K-means clustering** is a **data-driven** approach. Instead of relying on analyst-defined thresholds, it searches for groups of customers who are similar in their behaviour. This can reveal patterns that a fixed rule system might miss, but the results are often less immediately interpretable and depend on choices such as scaling and the number of clusters.

The main lesson is that both methods help us achieve the same broader goal: turning raw transaction data into customer-level insight. This is a good example of how descriptive analytics can move beyond simple summaries and support meaningful business decisions.

In practice:

* use **RFM** when you want a simple and transparent segmentation system
* use **k-means** when you want the data to help uncover behavioural groupings
* compare both when you want to understand the trade-off between business interpretability and data-driven discovery

