---
title: "Lecture 9: Instructor Notes"
author: "FBA Instructors"
format: html
---

## Learning Goals

* Explain why transaction-level data often need to be aggregated to the customer level for marketing decisions.
* Define **Recency**, **Frequency**, and **Monetary value** in customer analytics.
* Show how to build rule-based customer segments using RFM.
* Introduce k-means clustering as a data-driven alternative.
* Compare rule-based and data-driven segmentation.

## The Business Challenge: From Transactions to Customers

### What to say

* We observe **transactions**, but the manager wants to act on **customers**.
* A loyalty campaign is not targeted at rows in a spreadsheet. It is targeted at people.
* So the first analytical move is to aggregate purchases into customer-level summaries.
* This is a good example of descriptive analytics doing more than summarising the past.
* Use the line: **descriptive analytics is not just a summary of the past, but also a way to organise the past into forms that are useful for business action.**
* Today we will compare two ways of doing that:

  * **RFM**, which is rule-based
  * **k-means clustering**, which is data-driven

## RFM Segmentation in Practice

### What to say

* In marketing and CRM, a common way to summarise purchase behaviour is **RFM**.
* **Recency** = how recently the customer purchased.
* **Frequency** = how often the customer purchased.
* **Monetary value** = how much the customer spent.
* Why firms like it:

  * simple
  * interpretable
  * easy to connect to action
* Examples:

  * recent, frequent, high-spending customers may be good loyalty targets
  * formerly valuable but inactive customers may be reactivation targets
* Important framing:

  * RFM is not magic
  * it is a practical scoring framework built from data firms often already have
* Transition:

  * before we can segment customers, we need to turn transaction data into one row per customer

## Building Rule-Based Customer Segments

### Load packages and data

```{r}
library(tidyverse)
library(lubridate)
library(recipes)
library(factoextra)
library(broom)
```

```{r}
purchases <- read_csv("../data/cdnow_purchases.csv")

glimpse(purchases)
```

### What to say

* Each row is a **purchase**, not a customer.
* We need to move from transaction-level data to customer-level data.
* Encourage students to inspect the columns before doing anything.

### Aggregate to the customer level

```{r}
customer_aggregates <-
  purchases |>
  group_by(users_id) |>
  summarise(
    last_purch_date = max(date),
    frequency       = n(),
    monetary        = sum(amt),
    .groups = "drop"
  )
```

### What to say

* `group_by(users_id)` says: do the next calculations separately for each customer.
* `max(date)` gives the most recent purchase date.
* `n()` counts how many purchases they made.
* `sum(amt)` adds up their total spending.
* We are starting to build customer profiles from purchase histories.

### Compute recency

```{r}
analysis_date <- as_date("1998-07-01")

customer_aggregates <-
  customer_aggregates |>
  mutate(
    recency = interval(last_purch_date, analysis_date) %/% months(1)
  ) |>
  select(-last_purch_date)

glimpse(customer_aggregates)
```

### What to say

* Recency needs a reference date.
* We do **not** use today because these are historical data.
* We choose an analysis date that makes sense in the context of the dataset.
* After this step, each row is a customer with three useful summaries: recency, frequency, monetary.

### Turn raw measures into RFM scores

```{r}
rfm_data <-
  customer_aggregates |>
  mutate(
    r_tile = ntile(recency, 5),
    f_tile = ntile(frequency, 5),
    m_tile = ntile(monetary, 5)
  )

glimpse(rfm_data)
```

### What to say

* `ntile(..., 5)` splits each variable into five ranked groups.
* This is a simple scoring system.
* It makes customers easier to compare than using raw values alone.

### Flip recency so higher is better

```{r}
rfm_data <-
  rfm_data |>
  mutate(
    r_tile = ntile(-recency, 5)
  )

rfm_data |>
  group_by(r_tile) |>
  summarise(
    mean_r = mean(recency),
    .groups = "drop"
  )
```

### What to say

* Small recency means more recent purchasing.
* But for scoring, we often want high numbers to mean more desirable behaviour.
* So we reverse the scale.
* The summary check is good practice: verify that higher `r_tile` now corresponds to lower average recency.

### Assign business-defined segments

```{r}
rfm_segmented <-
  rfm_data |>
  mutate(
    segment = case_when(
      r_tile == 5 & f_tile == 5 & m_tile == 5 ~ "Champions",
      r_tile %in% 3:5 & f_tile %in% 3:5 & m_tile %in% 2:5 ~ "Potential Loyalist",
      r_tile %in% 2:4 & f_tile %in% 2:4 & m_tile %in% 2:4 ~ "Loyal Customers",
      r_tile %in% 3:4 & f_tile %in% 1:3 & m_tile %in% 3:5 ~ "Promising",
      r_tile %in% 4:5 & f_tile %in% 1:3 & m_tile %in% 1:5 ~ "New Customers",
      r_tile %in% 1:2 & f_tile %in% 3:4 & m_tile %in% 4:5 ~ "Can't Lose Them",
      r_tile %in% 1:2 & f_tile %in% 2:5 & m_tile %in% 4:5 ~ "At Risk",
      r_tile %in% 1:3 & f_tile %in% 3:5 & m_tile %in% 3:5 ~ "Need Attention",
      r_tile %in% 2:3 & f_tile %in% 1:3 & m_tile %in% 1:4 ~ "About To Sleep",
      r_tile == 1 & f_tile %in% 1:5 & m_tile %in% 1:5 ~ "Lost",
      .default = "Unclassified"
    )
  )
```

### What to say

* Do **not** teach this as syntax to memorise.
* Teach it as an example of how firms translate scores into operational labels.
* These labels make segmentation easier to communicate to managers.
* But the thresholds are chosen by analysts, not discovered by the data.

### Summarise the segments

```{r}
rfm_segmented |>
  group_by(segment) |>
  summarise(
    avg_recency   = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary  = mean(monetary),
    n_customers   = n(),
    .groups = "drop"
  ) |>
  arrange(desc(n_customers)) |>
  kableExtra::kable(digits = 2)
```

### What to say

* This is where the segmentation becomes interpretable.
* Ask: what kind of customer does each group seem to represent?
* Key strength of RFM:

  * easy to explain
  * easy to implement
  * easy to connect to action
* Transition:

  * but this approach depends on cutoffs that we chose ourselves

## A Data-Driven Alternative: Clustering Customers

### What to say

* Now ask a different question: what if we let the data suggest the groups?
* K-means clustering groups customers based on similarity.
* It is still segmentation, but the grouping rule comes from the algorithm rather than from predefined thresholds.

### Check ranges of the variables

```{r}
rfm_data |>
  summarise(
    min_recency   = min(recency),
    max_recency   = max(recency),
    min_frequency = min(frequency),
    max_frequency = max(frequency),
    min_monetary  = min(monetary),
    max_monetary  = max(monetary)
  ) |>
  pivot_longer(
    everything(),
    names_to = c("stat", "variable"),
    names_sep = "_"
  ) |>
  pivot_wider(
    names_from = stat,
    values_from = value
  )
```

### What to say

* The variables are on different scales.
* K-means uses distance.
* If one variable has much bigger numbers, it can dominate the clustering.
* So before clustering, we scale the variables.

### Scale the data with `recipes`

```{r}
rfm_rec <-
  recipe(~ recency + frequency + monetary, data = rfm_data) |>
  step_normalize(all_numeric_predictors())

rfm_prepped <- prep(rfm_rec)
rfm_scaled  <- bake(rfm_prepped, new_data = NULL)

head(rfm_scaled)
```

### What to say

* Be very explicit here.
* `recipe(...)` defines a preprocessing plan.
* At this point, nothing has been done to the data yet.
* It just records the steps we want to apply.
* `step_normalize(...)` adds a step to that plan.
* Here the step is: standardise each numeric predictor so it has mean 0 and standard deviation 1.
* `prep(rfm_rec)` estimates what is needed to carry out that plan using the data we supplied.
* In this case, it learns the means and standard deviations needed for normalisation.
* `bake(rfm_prepped, new_data = NULL)` actually applies the prepared steps to the data.
* So the logic is:

  * define the recipe
  * prepare the recipe
  * bake the transformed data
* This is useful because the same preprocessing pipeline can be reused consistently later.

### Verify scaling

```{r}
rfm_scaled |>
  summarise(
    across(everything(), list(mean = mean, sd = sd))
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_sep = "_"
  ) |>
  kableExtra::kable(digits = 2)
```

### What to say

* Always check what your preprocessing did.
* After scaling, each variable should have mean about 0 and standard deviation about 1.

### Fit k-means

```{r}
set.seed(123)

kmeans_3 <- kmeans(rfm_scaled, centers = 3)
```

### What to say

* We set a seed for reproducibility.
* `centers = 3` tells the algorithm to form three clusters.
* For today, we treat the choice of 3 as given.

### Visualise clusters

```{r}
fviz_cluster(
  kmeans_3,
  data = rfm_scaled,
  geom = "point",
  palette = "jco",
  ggtheme = theme_minimal()
)
```

### What to say

* The visualisation helps us see whether the groups look separated.
* Remind students that the plot is based on the scaled inputs.
* The important question is not just whether the plot looks nice, but whether the clusters seem meaningfully different.

### Attach cluster labels back to the data

```{r}
rfm_clustered <-
  augment(kmeans_3, rfm_data)

head(rfm_clustered)
```

```{r}
rfm_clustered |>
  group_by(.cluster) |>
  summarise(
    avg_recency   = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary  = mean(monetary),
    n_customers   = n(),
    .groups = "drop"
  ) |>
  arrange(.cluster) |>
  kableExtra::kable(digits = 2)
```

### What to say

* `augment()` attaches the model output back to the data.
* This is what makes the clusters interpretable.
* Without going back to the original variables, cluster labels are just numbers.
* Ask: which cluster seems to contain more valuable customers? Which seems more at risk?

### If time: choosing the number of clusters

```{r}
set.seed(42)

rfm_scale_sample <-
  rfm_scaled |>
  sample_n(1000)
```

```{r}
fviz_nbclust(
  rfm_scale_sample,
  kmeans,
  method = "wss",
  k.max = 15,
  nstart = 5
)
```

### What to say

* Only do this if time allows.
* The point is that the number of clusters is a modeling choice.
* The algorithm does not automatically know the “correct” number.

## Comparing the Two Approaches

### What to say

* RFM is **rule-based**.

  * transparent
  * manager-friendly
  * easy to connect to action
* K-means is **data-driven**.

  * can reveal patterns we did not specify in advance
  * but needs preprocessing
  * and can be harder to interpret immediately
* Main lesson:

  * same transaction histories
  * two different segmentation logics
  * both help turn past behaviour into forms useful for business decisions
* Nice final line:

  * the data have not changed, but the way we organise them changes what the business can see and do



